Hybrid Threads for the Parrot Virtual Machine

### 2012-02-27 Paper.tex
First feedback

\AddBibFile{literature.bib}

@@

On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a very flexible and widely used programming language but had started to show its age and suffered from early design decisions. The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility\cite{Masak40451}. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions\cite{Developer}. Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot\cite{Perl}.

@@

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multi-threading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

\section{Why is multi-threading support so important?}

In the year 2005 development of faster CPUs shifted from increased speed of a single core to adding more cores. Modern processors contain up to 12 cores with even mobile phones having up to four. To utilize a modern CPU's power, code needs to be run in parallel. In UNIX (and thus Perl) tradition, this is accomplished using multiple processes being a good solution for many use cases. For many others like Perl 6's auto threading of hyper operators, the cost of process setup and communication would be prohibitively high except for very large data sets.

\section{Why is it so difficult to implement multi-threading support in Perl or Parrot?}

@@

While there are multi threaded garbage collection schemes, their implementation has not yet been attempted in Parrot.

@@

This paper suggests a new approach based on a hybrid threading system. So called green threads are used to simplify the implementation of a nearly lock free multi threading implementation. This approach is based on a design by Andrew Whitworth and Nat Tuck.

@@

In Java responsibility for preventing concurrency issues is obliged on the user code. The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself. The class library provides the user with high level data structures explicitely designed for multi threaded scenarios.

Java version 1.1 used ``green threads'' to support multithreaded execution of Java programs. Green threads are threads simulated by the virtual machine but unable to use more than one CPU core for processing. Details are described in chapter~\ref{cha:green_threads}. Version 1.2 introduced native OS thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitely type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multi threading.

@@

Parrot consists of the virtual machine (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit). This paper concentrates on the virtual machine itself. The interpreter itself is written in C. Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.

@@

PMCs are containers for all higher level types such as objects, arrays, hash tables or code. Thus they are very similar to Python's PythonObject types. PMCs are defined by C structs and are fully garbage collected. Their definition looks like:
\begin{CCode}
struct PMC {
    Parrot_UInt    flags;
    VTABLE         *vtable;             /* Pointer to vtable. */
    DPOINTER       *data;               /* Pointer to attribute structure. */
    PMC            *_metadata;          /* Pointer to metadata PMC. */
};
\end{CCode}
In short it contains a pointer to a type specific data structure and some meta data. The vtable is where the type's behaviour is defined. It contains a long list of function pointers forming Parrot's unified type interface. This interface is a union of numeric, string, array, hash and object like behaviours. For example the  \textit{get\_integer} function returns an integer value for the data type. For a simple \textit{int} it is the contained value. For an array it may be the number of elements. The \textit{find\_method} function makes user defined methods of objects possible.

@@

\section{Garbage collector}

Parrot supports different garbage collector implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
\begin{itemize}
\item Inf: a GC for debugging purposes never collecting any garbage
\item MS: a basic mark and sweep implementation
\item MS2: a non-recursive mark and sweep implementation
\item GMS: a generational, non-compacting, mark and sweep GC
\end{itemize}
with GMS being the default.

@@

Let's look at a simple example to explain why concurrent processing can endanger the interpreter's internal consistency. The basic problem is concurrent writes to shared data. Assumeing an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
\begin{CCode}
pmclass ResizableIntegerArray auto_attrs provides array {
    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
    ATTR INTVAL * int_array; /* INTVALs are stored here */
\end{CCode}
to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
Now if two threads simultainously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size. In this case one of the appended values would be lost.

@@

The scheduler is implemented as a PMC (Parrot's polymorphic container type). The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy. Features a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.

@@

In keeping the analogy of the interpreter being a software CPU, multi threading is implemented by having a separate interpreter with its own register set for each thread. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active. Other VMs solve this problem by suspending all threads while the GC is running. While multi threaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.

>>>>

\usepackage{wrapfig}
\AddBibFile{literature.bib}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
A simple example of Perl:
\include{perlexample}
\end{wrapfigure}
%
On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a very flexible and widely used programming language but had started to show its age and suffered from early design decisions. The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility\cite{Masak40451}. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions\cite{Developer}. Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot\cite{Perl}.

@@

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

\section{Why is multithreading support so important?}

In the year 2005 development of faster CPUs (Central Processing Units) shifted from increased speed of a single core to adding more cores. Modern processors contain up to 12 cores with even mobile phones having up to four. To utilize a modern CPU's power, code needs to be run in parallel. In UNIX (and thus Perl) tradition, this is accomplished using multiple processes being a good solution for many use cases. For many others like Perl 6's auto threading of hyper operators, the cost of process setup and communication would be prohibitively high except for very large data sets.

\section{Why is multithreading support so difficult to implement?}

@@

@@

This paper suggests a new approach based on a hybrid threading system. So called green threads are used to simplify the implementation of a nearly lock free multithreading implementation. This approach is based on a design by Andrew Whitworth and Nat Tuck. The goal of this paper is to demonstrate the advantages of this model and produce a working implementation which can be used to investigate the performance characteristics of a hybrid threading system.

@@

In Java responsibility for preventing concurrency issues is obliged on the user code. The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself. The class library provides the user with high level data structures explicitely designed for multithreaded scenarios.

Java version 1.1 used ``green threads'' to support multithreaded execution of Java programs. Green threads are threads simulated by the virtual machine but unable to use more than one CPU core for processing. Details are described in chapter~\ref{cha:green_threads}. Version 1.2 introduced native OS (Operating System) thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitely type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.

@@

Parrot consists of the virtual machine (VM, also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit). This paper concentrates on the virtual machine itself. The interpreter itself is written in C. Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.

@@

PMCs (PolyMorphic Container) are containers for all higher level types such as objects, arrays, hash tables or code. Thus they are very similar to Python's PythonObject types. PMCs are defined by C structs and are fully garbage collected. Their definition looks like:
\begin{CCode}
struct PMC {
    Parrot_UInt    flags;
    VTABLE         *vtable;             /* Pointer to vtable. */
    DPOINTER       *data;               /* Pointer to attribute structure. */
    PMC            *_metadata;          /* Pointer to metadata PMC. */
};
\end{CCode}
In short it contains a pointer to a type specific data structure and some meta data. The vtable is where the type's behaviour is defined. It contains a long list of function pointers forming Parrot's unified type interface. This interface is a union of numeric, string, array, hash and object like behaviours. For example the  \textit{get\_integer} function returns an integer value for the data type. For a simple \textit{int} it is the contained value. For an array it may be the number of elements. The \textit{find\_method} function makes user defined methods of objects possible.

@@

\section{Garbage Collector}

Parrot supports different GC (Garbage Collector) implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
\begin{itemize}
\item Inf: a GC for debugging purposes never collecting any garbage
\item MS: a basic mark and sweep implementation
\item MS2: a non-recursive mark and sweep implementation
\item GMS: a generational, non-compacting, mark and sweep GC
\end{itemize}
with GMS being the default.

@@

Following is a simple example to explain why concurrent processing can endanger the interpreter's internal consistency. The basic problem is concurrent writes to shared data. Assumeing an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
\begin{CCode}
pmclass ResizableIntegerArray auto_attrs provides array {
    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
    ATTR INTVAL * int_array; /* INTVALs are stored here */
\end{CCode}
to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
Now if two threads simultainously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size. In this case one of the appended values would be lost.

@@

The scheduler is implemented as a PMC (Parrot's polymorphic container) type. The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy. Features a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.

@@

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active. Other VMs solve this problem by suspending all threads while the GC is running. While multithreaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.
